HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations

### 2021-02-20 docs/main.tex
introduced HAM enrichment plot

Consider a phylogenetic tree \(\Psi\) obtained from standard maximum likelihood methods for a given multiple sequence alignment. We are interested in estimating \(P(y_{ik}=1|\Psi)\), that is, the probability of observing the replacement \(k\) in sequence \(i\) based on the underlying phylogenetic model. A quantity that can be readily computed using phylogenetic software like RAxML-NG~\cite{Kozlov2019} is \(P(\Psi|y_{ik}=1)\). For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y_{ik}\) at its leaves and optimize the branch lengths. \(P(\Psi|y_{ik}=1)\) is then the likelihood of the annotated phylogenetic tree. Similarly, we can also compute \(P(\Psi|y_{ik}=0)\) by flipping the annotation of sequence $i$ from 1 to 0 (keeping all other observations). With  $P(\Psi|y_{ik}=1)$ and $P(\Psi|y_{ik}=0)$ known and the relative frequencies of 0 an 1 as priors, we can estimate $P(y_{ik}=1|\Psi)$ by applying Bayes' theorem.

In order to check whether the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see Fig.~\ref{fig:phylogeny-calibration}). All observations are binned according to their inferred probability \(P(y_{ik}=1|\Psi)\). If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ik}=1|\Psi)\) of around x\% we really do observe \(y_{ik}=1\) around x\% percent of the time.

The estimated probabilities based on phylogeny are then included in the model as additional intercepts (second term of logistic argument):

@@

\subsection{Model implementation}
All models were implemented in Stan~\citep{Stan2021}. The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
A Julia~\citep{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data. Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux. %We additionally provide Docker images to easily run the model on any operating system in a containerized environment, assuming Docker is installed.

@@

@@

  Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.

  Here, we use these so-called posterior predictive checks to compare the calibration of the predicted probabilities of observing the replacement to the actual observed data.

  \subsubsection{Convergence diagnostics}

  We use Stan (version 2.23)~\cite{Stan2021} to specify and fit our Bayesian models.
  Stan provides an implementation of a dynamic Hamilitonian Monte Carlo sampler, which is a powerful technique to approximate expectations with respect to the posterior distribution~\cite{Betancourt2017}. One advantage of using such methods is that they provide sensitive diagnostics to identify a wide range of possible sampling issues.
  We use the split-\(\hat{R}\) convergence diagnostic~\cite{Gelman1992} to identify convergence issues. We require a value of \(\hat{R}\) below 1.01 for all model parameters. Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions. A detailed explanation of these quantities is provided in the Stan reference manual.

@@

@@

\subsubsection{Calibration}
 \begin{figure}[!ht]
  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
  \caption{Calibration plot for the HBV dataset. For a description of all datasets used in this study see section~\ref{sec:results}. Calibration plots for the other datasets are shown in the supplementary.
  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1|\Psi)\) and then grouped into \(n\) bins.
  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability). The error bars show the cutpoints for each bin. If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
  \label{fig:phylogeny-calibration}
\end{figure}

@@

\subsection*{Comparison to a list of known epitopes}

When testing models, we prefer to evaluate them on real-world data whenever possible. 
However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before. 
Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.

To circumvent these issues and still compare against real-world data, we chose the following strategy:
For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation. That is, for Bayesian models we calculate the probability of the corresponding regression coefficient exceeding zero \(P(\beta_{HLA} > 0)\), for Fisher's exact test we sort HLA allele - replacement pairs by increasing p-values and for the Phylogenetic Dependency Network, we rank the associations by increasing q-values.
Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs. the corresponding rank. The underlying assumption for this kind of model check is that we expect to see an enrichment of replacements that lie inside the region of known epitopes and that this enrichment is stronger for better performing models.

\begin{figure}[ht!]
  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
  \caption{Number of associations inside the boundary of known epitopes vs. rank. For each model, the pairs of HLA alleles and replacements are sorted by decreasing confidence of being an HLA associated mutation. Then, each replacement is compared to a list of known epitopes and a line of the cumulative number of associations inside known epitopes vs. the current rank is plotted for each model. Better performing models are expected to have an enrichment of associations that are inside of known epitopes. 
  PhyloD: Phylogenetic Dependency Network, Model 1: simple logistic regression model with broad Student-t priors, Model 2: logistic regression model with horseshoe prior, Model 3: logistic regression model with horseshoe prior and phylogeny, Model 4: Full model with epitope prediction. The thin gray lines denote random permutations of the list of HLA allele - replacement pairs and act as a baseline.}
  \label{fig:comparison}
\end{figure}

>>>>

Consider a phylogenetic tree \(\Psi\) obtained from standard maximum likelihood methods for a given multiple sequence alignment. We are interested in estimating \(P(y_{ik}=1|\Psi)\), that is, the probability of observing the replacement \(k\) in sequence \(i\) based on the underlying phylogenetic model. A quantity that can be readily computed using phylogenetic software like RAxML-NG~\cite{Kozlov2019} is \(P(\Psi|y_{ik}=1)\). For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y_{ik}\) at its leaves and optimize the branch lengths. \(P(\Psi|y_{ik}=1)\) is then the likelihood of the annotated phylogenetic tree. Similarly, we can also compute \(P(\Psi|y_{ik}=0)\) by flipping the annotation of sequence $i$ from 1 to 0 (keeping all other observations). With  $P(\Psi|y_{ik}=1)$ and $P(\Psi|y_{ik}=0)$ known and the relative frequencies of 0 an 1 as priors, we can estimate $P(y_{ik}=1|\Psi)$ by applying Bayes' theorem. The estimated probabilities based on phylogeny are then included in the model as additional intercepts (second term of logistic argument):

@@

\subsection{Model implementation}
All models were implemented in Stan 2.23 \citep{Stan2021}, a probabilistic programming language and Hamiltonian Monte Carlo sampler for efficient numerical computation of posterior distributions. The Stan code is available \hl{DaHo: at ...} in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities. A Julia \citep{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data. Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux. %We additionally provide Docker images to easily run the model on any operating system in a containerized environment, assuming Docker is installed.

@@

\subsubsection{Convergence diagnostics}
We use the split-\(\hat{R}\) convergence diagnostic to identify Markov chain convergence issues \citep{Gelman1992, Stan2021}. We require a value of \(\hat{R}\) below 1.01 for all model parameters. Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions \citep{Stan2021}.

@@

In posterior predictive checks, we simulate new data from the inferred posterior distribution and the likelihood, and we compare these simulated data with representative real data \citep{Gabry2019}. A good model should predict data that are consistent with real data. This general idea was employed in two ways to test our models.

First, we checked with \emph{calibration plots} whether the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data (Fig.~\ref{fig:calibration}). All observations were binned according to their inferred probability \(P(y_{ik}=1|\Psi)\). If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ik}=1|\Psi)\) of around x\% we really do observe \(y_{ik}=1\) around x\% percent of the time.

Second, we assessed the abilities of different models and methods to discover HAMs with \emph{HAM enrichment plots}. These plots are based on the observation that CTL escape mutations are enriched in epitopes \citep{Bronke2013}. Hence, the degree by which methods for HAM prediction recover this trend is a measure of model performance. To implement this measure, we first ranked all evaluated replacements according to their respective credibility of being a HAM, computed as integral of the marginal posterior $P(\beta_{HLA}>0)$. For comparison with established methods, namely Fisher's exact test and Phylogenetic Dependency Network \citep{Carlson2008}, ranked lists based on p- and q-values, respectively, were computed. Then we calculated for each rank $r$ the accumulated number $N_e(r)$ of predictions of this rank or better ranks were located inside known epitopes. The higher the curve $N_e(r)$, the higher the enrichment of predicted HAMs in epitopes, see e.g.\ Fig.\ \ref{fig:comparison}.

@@

\hl{DaHo: sources of known epitopes?}

@@

\subsubsection{Calibration}

@@

\subsection{Comparison to a list of known epitopes}

%However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before. 
%Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
%A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.

%To circumvent these issues and still compare against real-world data, we chose the following strategy:
%For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation. That is, for Bayesian models we calculate the probability of the corresponding regression coefficient exceeding zero \(P(\beta_{HLA} > 0)\), for Fisher's exact test we sort HLA allele - replacement pairs by increasing p-values and for the Phylogenetic Dependency Network, we rank the associations by increasing q-values.
%Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs. the corresponding rank. The underlying assumption for this kind of model check is that we expect to see an enrichment of replacements that lie inside the region of known epitopes and that this enrichment is stronger for better performing models.

\begin{figure}[ht!]
  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
  \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$. PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction. The thin gray line are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baseline.}
  \label{fig:comparison}
\end{figure}
